Refinements for open automata

### 2022-01-03 2021-Refinement/paper/refinementarticle.tex
Proof of lemm 2 in progress

Before characterising a composition which does not introduce deadlocks we need to be able to tell which states are reachable.
\begin{definition}[Reachability]
For any open automata \(A = \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is any predicate on states that is valid on initial state, and preserved across transitions:
\[\sigma_0 \vdash \reach{A}\mpar{s_0}\quad\land\quad\forall t = \OTg \in T, \fvars{t} \vdash \left(\reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}\right)\]
\end{definition}
Reachability takes into account all paths, and can over-approximate the reachable configurations. 
\TODO{rediger la phrase suivante}
Because of the over approximation, to implement a verifier it is possible to try find a formula that over-approximate the reachability property and is representable in the tool.
% LUDO HERE
\begin{definition}[Non-locking composition]
Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.

@@

- composition is a refinement (under which condition? non-locking composition?)

@@

\section{Proof of Context Refinement}
Let us denote by $A_{13}$ (resp. $A_{23}$) the open automaton resulting from $A_1\subst{A_3}{k}$ ($A_2\subst{A_3}{k}$),  to prove the theorem it sufficient to prove that there exists a relation between states of the two open automata that satisfies the conditions of the Definition \ref{Def:OA-Refinement}. Let $R$ be the refinement relation relating states of $A_1$ and $A_2$.\\
Let us consider any two states $s_{13} = \mpar{s_1,s_3}$ and $s_{23} = \mpar{s_2,s_3'}$ we define a relation $R'$ relating states of $A_{13}$ and $A_{23}$ as follows:
\[R' \mpar{s_{13}, s_{23}} = \choice{R\mpar{s_{1}, s_{2}} \wedge \reach{A_{13}}\mpar{s_{13}} \wedge\!\!\!\! \bigwedge_\subbox{v\in \fvars{s_{3}}
}\!\!\! \sigma_{13}(v)=\sigma_{23}(v) & \text{ if } s_{3} = s'_{3} \\ \bot & \text{otherwise}} \]
We want to prove that $\mpar{R', H\cup J_3 \setminus\{k\}}$ is a hole-tracking simulation of $A_{13}$ and $A_{23}$.
\begin{enumerate}
\item By using the fact $R$ relates initial configurations of  $A_1$ and $A_2$, initial states are reachable and by renaming $\sigma_{03}$ by  $\sigma_{013}$ and  $\sigma_{023}$ we need to prove:
\[\sigma_{013} \uplus \sigma_{023} \vdash R'\mpar{s_{013}, s_{023}}\]

We have by hypothesis:
\begin{align*}
& \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge 
\mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}\\
\iff & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}} \wedge \reach{A_{13}}\mpar{s_{013}} \psubst{\sigma_{013}} \wedge \bigwedge_\subbox{v\in V_3} \sigma_{03}(v)=\sigma_{03}(v)\\
\iff & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}} \wedge \reach{A_{13}}\mpar{s_{013}} \psubst{\sigma_{013}} \wedge \bigwedge_\subbox{v\in V_3} \sigma_{013}(v)=\sigma_{023}(v)\\
\iff & R\mpar{s_{01}, s_{02}} \wedge \reach{A_{13}}\mpar{s_{013}}  \wedge \bigwedge_\subbox{v\in V_3} \sigma_{013}(v)=\sigma_{023}(v) \psubst{ \sigma_{01} \uplus \sigma_{02} \uplus \sigma_{013} \uplus \sigma_{023}}
\end{align*}
\item We need to prove for any OT $t_{13}$ originating from $s_{13}$:
\begin{align*}
		&  \OTx{13}{}{13}{13} \in \fOT{s_{13}}
\end{align*}		
there exist an indexed family $t_{23x}$ of OTs originating from $s_{23}$: 
\begin{align*}
		&  \mpar{\OTx{23}{x}{23x}{23x} \in \fOT{s_{23}}}^{x \in X}
\end{align*}	
\begin{align*}		
		& \quad \mpar{\forall x \in X, J'_{23x} \cap H = J'_{13} \cap H} \nwedge\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash R\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H} \beta_{13j} = \beta_{23xj} \\[12pt]
			\nwedge g_{23x} \wedge R\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}
According to Definition \ref{Def:CompOA}, we know that $t_{13}$ is built by:
 \[  \choice{
	\OTx{1}{}{1}{1} \wedge  \OTx{3}{}{3}{3}
	& \text{if } k \in J'_{1} \\
	\OTx{1}{}{1}{1}	 & \text{otherwise}
	} \]

By hypothesis, we have for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exist an indexed family of OTs originating from $s_{2}$: 
\begin{align*}
\mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_{2}}}^{x \in X} 
\end{align*}
such that $\forall x \in X, J'_{2x} \cap H = J'_1 \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\ \mpar{R\mpar{s_1, s_{2}} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} 
\end{multline*}

@@

\[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\].

>>>>

Before characterising a composition which does not introduce deadlocks we need to be able to tell which states are reachable.
\begin{definition}[Reachability] \label{Def:Reach}
For any open automata \(A = \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is any predicate on states that is valid on initial state, and preserved across transitions:
\[\sigma_0 \vdash \reach{A}\mpar{s_0}\quad\land\quad\forall t = \OTg \in T, \fvars{t} \vdash \left(\reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}\right)\]
\end{definition}
Reachability takes into account all paths, and can over-approximate the reachable configurations. 
\TODO{rediger la phrase suivante}
Because of the over approximation, to implement a verifier it is possible to try find a formula that over-approximate the reachability property and is representable in the tool.
% LUDO HERE
\begin{definition}[Non-locking composition]
Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.

@@

\begin{theorem}[Congruence]
Let $A_1$, $A_2$ and $A_3$ be three open automata with $\wrel{A_2}{A_3}{H}$. 
Suppose that \(k \in H\) and that \(A_1\subst{A_2}{k}\) is non-locking.
We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\].
\end{theorem}
- composition is a refinement (under which condition? non-locking composition?)

@@

\section{Proof of Context Refinement}
Let us denote by $A_{13}$ (resp. $A_{23}$) the open automaton resulting from $A_1\subst{A_3}{k}$ ($A_2\subst{A_3}{k}$),  to prove the theorem it sufficient to prove that there exists a relation between states of the two open automata that satisfies the conditions of the Definition \ref{Def:OA-Refinement}. Let $R$ be the refinement relation relating states of $A_1$ and $A_2$.\\
Let us consider any two states $s_{13} = \mpar{s_1,s_3}$ and $s_{23} = \mpar{s_2,q_3}$ we define a relation $R'$ relating states of $A_{13}$ and $A_{23}$ as follows:
\[R' \mpar{s_{13}, s_{23}} = \choice{R\mpar{s_{1}, s_{2}} \wedge \reach{A_{13}}\mpar{s_{13}} \wedge\!\!\!\! \bigwedge_\subbox{v\in \fvars{s_{3}}
}\!\!\! \sigma_{13}(v)=\sigma_{23}(v) & \text{ if } s_{3} = q_{3} \\ \bot & \text{otherwise}} \]
We want to prove that $\mpar{R', H\cup J_3 \setminus\{k\}}$ is a hole-tracking simulation of $A_{13}$ and $A_{23}$.
\begin{enumerate}
\item By using the fact $R$ relates initial configurations of  $A_1$ and $A_2$, initial states are reachable and by renaming $\sigma_{03}$ by  $\sigma_{013}$ and  $\sigma_{023}$ we need to prove:
\[\sigma_{013} \uplus \sigma_{023} \vdash R'\mpar{s_{013}, s_{023}}\]

We have by hypothesis:
\begin{align*}
& \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge 
\mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}\\
\iff & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}} \wedge \reach{A_{13}}\mpar{s_{013}} \psubst{\sigma_{013}} \wedge \bigwedge_\subbox{v\in V_3} \sigma_{03}(v)=\sigma_{03}(v)\\
\iff & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}} \wedge \reach{A_{13}}\mpar{s_{013}} \psubst{\sigma_{013}} \wedge \bigwedge_\subbox{v\in V_3} \sigma_{013}(v)=\sigma_{023}(v)\\
\iff & R\mpar{s_{01}, s_{02}} \wedge \reach{A_{13}}\mpar{s_{013}}  \wedge \bigwedge_\subbox{v\in V_3} \sigma_{013}(v)=\sigma_{023}(v) \psubst{ \sigma_{01} \uplus \sigma_{02} \uplus \sigma_{013} \uplus \sigma_{023}}
\end{align*}
\item We need to prove for any OT $t_{13}$ in $T_{13}$ originating from $s_{13}$:
\begin{align*}
		&  \OTx{13}{}{13}{13} \in \fOT{s_{13}}
\end{align*}		
there exists an indexed family $t_{23x}$ of OTs originating from $s_{23}$: 
\begin{align*}
		&  \mpar{\OTx{23}{x}{23x}{23x} \in \fOT{s_{23}}}^{x \in X}
\end{align*}	
\begin{align*}		
		& \text{such that } \mpar{\forall x \in X, J'_{23x} \cap H' = J'_{13} \cap H'} \text{ and }\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash R'\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H'} \beta_{13j} = \beta_{23xj} \\[12pt]
			\nwedge g_{23x} \wedge R'\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}
where 
\begin{align*}
&V_{13}= V_1\uplus V_3 \text{ and }
V_{23}= V_2\uplus V_3\\
& H'=J_3'\uplus H \setminus \{k\} \subseteq J_3'\uplus \mpar{J_1' \cap J_2'} \setminus \{k\} = \mpar{J_3'\uplus J_1' \setminus \{k\} } \cap \mpar{J_3'\uplus J_2' \setminus \{k\}} = J_{13}'\cap J_{12}'\\
\end{align*}
The open transition $t_{13}$ is obtained by the composition of $t_1=\OTx{1}{}{1}{1}$ and $t_3=\OTx{3}{}{3}{3} $, which gives,  according to Definition \ref{Def:CompOA}, an open transition in one of the following forms:
% \[  \choice{
%	\OTx{1}{}{1}{1} \text{ and }  \OTx{3}{}{3}{3}
%	& \text{if } k \in J'_{1} \\
%	\OTx{1}{}{1}{1}	 & \text{otherwise}
%	} \]
\[
\choice{
\OT{\mpar{s_1, s_3}}{\mpar{s'_1, s'_3}}{\alpha_1}{\beta_j^{j \in J'_1 \uplus J'_3 \setminus \{k\}}}{g_1 \wedge g_3 \wedge \alpha_3 = \beta_k}{\psi_1 \uplus \psi_3} & \text{ if } k \in J'_{1} \\
\OT{\mpar{s_1, s_3}}{\mpar{s'_1, s_3}}{\alpha_1}{\beta_j^{j \in J'_1}}{g_1}{\psi_1}& \text{otherwise}
}
\]

We have by hypothesis $\wrel{A_1}{A_2}{H}$, then for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{2}$: 
\begin{align*}
\mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_{2}}}^{x \in X} 
\end{align*}
such that $\forall x \in X, J'_{2x} \cap H = J'_1 \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\ \mpar{R\mpar{s_1, s_{2}} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}}  \qquad (*)
\end{multline*}	
We have  $k \in H$ and $k \in J'_{1}$ then  we have $k \in J'_{2x}$.\\
By composing the family of OTs  $t_{2x}^{x\in X}$ originating from  $s_2$ with an open transition $t_3 =\OT{q_3}{q_3'}{\alpha_{3}'}{\beta_j^{j \in J''_{3}}}{g_{3}'}{\psi_{3}'} \in T_3$ %${\OT{q_{3}}{q'_{3}}{\alpha_{3}}{\beta_{ j}^{j \in J'_{3}}}{g_{3}}{\psi_{3}}}$, 
gives a family of OTs $t_{23x}^{x\in X}$.  The result of the composition will be as follows:
\begin{enumerate}
\item if $ s_3=q_3$
\[
\choice{
\OT{\mpar{s_2, q_3}}{\mpar{s'_{2x}, q'_3}}{\alpha_{2x}}{\beta_j^{j \in J'_{2x} \uplus J''_3 \setminus \{k\}}}{g_{2x} \wedge g_3' \wedge \alpha_3' = \beta_{2xk}}{\psi_{2x} \uplus \psi_3'} & \text{ if } k \in J'_{2x} \\
\OT{\mpar{s_2, q_3}}{\mpar{s'_{2x}, q_3}}{\alpha_{2x}}{\beta_j^{j \in J'_{2x}}}{g_{2x}}{\psi_{2x}}& \text{otherwise}
}
\]
\item if $s_3 \neq q_3$ no element is produced that results an empty family.
\end{enumerate}

It remains to prove that we have:
\begin{align*}		
		&\mpar{\forall x \in X, J'_{23x} \cap H' = J'_{13} \cap H'} \text{ and }\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash R'\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H'} \beta_{13j} = \beta_{23xj} \\[12pt]
			\nwedge g_{23x} \wedge R'\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}

For all $x \in X$ we have two cases:
\begin{align*}
		\text{if } k \in J'_1: && J'_{23x} \cap H' & = \mpar{J'_{3} \uplus J'_{2x} \setminus \mbrc{k}} \cap  \mpar{J'_{3} \uplus H \setminus \mbrc{k}}\\
		&&& = J'_{3}  \uplus \mpar{J'_{2x} \cap H} \setminus \mbrc{k} \\
		&&& = J'_{3} \uplus \mpar{J'_1 \cap H} \setminus \mbrc{k} \\
		&&& = \mpar{ J'_{3} \uplus J'_1 \setminus \mbrc{k}} \cap \mpar{J'_{3} \uplus H \setminus \mbrc{k}}\\
		&&& = J'_{13} \cap H'\\
\text{if } k \notin J'_1: && J'_{23x} \cap H' & = \mpar{J'_{2x} \setminus \mbrc{k}} \cap \mpar{J'_3 \uplus H \setminus \mbrc{k}} \\
		&&& = \mpar{J'_{2x} \cap H} \setminus \mbrc{k} \\
		&&& = \mpar{J'_1 \cap H} \setminus \mbrc{k}  \\
		&&& = J'_3 \uplus \mpar{J'_1 \cap H} \setminus \mbrc{k}  \\
		&&& = J'_{13} \cap H'
		\end{align*}

By hypothesis $\reach{A_{13}}\mpar{s_{13}}$ holds,  we then have according to the Definition \ref{Def:Reach} the following:
\[ \fvars{t_{13}} \vdash \left(\reach{A_{13}}\mpar{s_{13}} \wedge g_{13} \implies \reach{A_{13}}\mpar{s_{13}'}\psubst{\psi_{13}}\right) \]	
\TODO{Un peu de gymnastique est necessaire sur les valuations, etc...}
From this formula and $(*)$ 
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus V_3 \uplus \fvars{t_1} \vdash\\ \mpar{R\mpar{s_1, s_{2}} \wedge \reach{A_{13}}\mpar{s_{13}} \wedge g_{13} 
\wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} \wedge \reach{A_{13}}\mpar{s_{13}'}\psubst{\psi_{13}}
\end{multline*}

\TODO{Pas termine: valider d'abord la definition de R}
\item The last formula to prove is the following:
\begin{multline*}
\forall \mpar{s_{13}, s_{23}} \in S_{13} \times S_{23},\\ V_{13} \uplus V_{23} \uplus \biguplus_\subbox{t_{23}\in \fOT{s_{23}}} \fvars{t_2} \vdash \left(R\mpar{s_{13}, s_{23}} \wedge \bigvee_\subbox{t_{23} \in \fOT{s_{23}}} \fguard{t_{23}} \implies \bigvee_\subbox{t_1 \in \fOT{s_{13}}} \fguard{t_{13}} \right)
\end{multline*}
By hypothesis we have the following:
\begin{multline*}
\forall \mpar{s_1, s_2} \in S_1 \times S_2,\\ V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash \left(R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \right)
\end{multline*}
By adding the term $\displaystyle{\bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3}}$ to both sides of the implication, we get: 
\begin{multline*}
\forall \mpar{s_1, s_2,s_3} \in S_1 \times S_2 \times S_3, V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash\\ \left(R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \wedge \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \wedge \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3}\right)
\end{multline*}
By hypothesis we also have \(A_1\subst{A_3}{k}\) is non-locking,  thus we can write the following:
\[ \fvars{t_{13}} \vdash \left(\reach{A_{13}}\mpar{s_{13}} \wedge g_{13} \implies \reach{A_{13}}\mpar{s_{13}'}\psubst{\psi_{13}}\right) \]

@@

%\[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\].
